Quantile Forecast Matching with a Bayesian Quantile Gaussian Process Model

### 2025-01-16 quantile_matching.tex
Rework images in manuscript

%% The amssymb package provides various useful mathematical symbols
% \usepackage{amssymb}
%% The amsmath package provides various useful equation environments.
% \usepackage{amsmath}
\usepackage{amsmath,amsthm,amssymb,scrextend,mathtools}
\usepackage{hyperref}
\usepackage{newtxmath} %for making indicator function
% \RequirePackage[colorlinks,citecolor=blue,urlcolor=blue,backref=page,backref=page]{hyperref}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\usepackage{natbib}

@@

%% Abstract
\begin{abstract}
For various reasons a relatively small number of probabilities along with corresponding quantiles are often used to define predictive distributions or probabilistic forecasts. These quantile predictions offer easily interpreted uncertainty quantification of an event, and quantiles are generally straightforward to estimate using standard statistical and machine learning methods. However, compared to a distribution defined by a probability density or cumulative distribution function, a set of quantiles has far less distributional information. When given estimated quantiles it may be desirable to estimate a fully defined continuous distribution function, and many researchers do so to make evaluation or ensemble modeling simpler. Most existing methods for fitting a distribution to quantiles lack accurate representation of the inherent uncertainty from quantile estimation or are limited in their applications. In this manuscript we present a Gaussian process model, the quantile Gaussian process, which is based on established theory of quantile functions and sample quantiles, to construct a probability distribution given estimated quantiles. A Bayesian application of the quantile Gaussian process is evaluated for parameter inference and distribution approximation in simulation studies, and the quantile Gaussian process is used to approximate the distributions of quantile forecasts from the 2023-24 US Centers for Disease Control collaborative flu forecasting competition. The simulation studies and data analysis show that the quantile Gaussian process leads to accurate inference on model parameters, estimation of a continuous distribution, and uncertainty quantification of sample quantiles.
\end{abstract}

@@

The use of quantiles in statistical modeling and in reporting inferential or predictive uncertainty is widespread, and reporting several quantiles or predictive intervals is common for probabilistic forecasting \cite[]{gneiting2023model}. Quantiles can be easier to interpret than statistical model parameters and are often used to define confidence or prediction intervals. Thus with multiple quantiles for a particular outcome, one has a measure of uncertainty.
Estimating quantiles in the presence of covariates via 
quantile regression is a common
statistical and machine learning strategy, and where parametric models are complicated or nonexistent quantiles may be easier to estimate via machine learning methods \cite[]{martin2022direct,chung2021beyond, koenker2017quantile, koenker1978regression}. In estimating multiple quantiles, quantile regression often provides a tradeoff with parametric modeling in that for some problems quantile regression may be easier to perform but that predicted quantiles lack the detailed information of a fully defined predictive distribution \cite[]{pohle2020murphy}. Quantile regression may also lead to other issues such as quantile crossing where an estimated quantile may be smaller than another estimated quantile with for a lower probability level \cite[]{chernozhukov2010quantile,he1997quantile}.
% Data is often reported as quantiles, 
Perhaps, for data privacy reasons, data quantiles are often reported as summaries of the data.
% along with predictive distributions. 
% \jarad{Predictive distributions for what?}
Census or medical data, which can be very large or personal to the subjects, may be published as summary or aggregate data including percentiles (quantiles) and medians \cite[]{simpson2023interpolating,cdc2022growthcharts,nirwan2020bayesian}. In collaborative forecast initiatives and competitions, probabilistic forecasts are often submitted as a set of predictive quantiles for given probabilities \cite[]{gneiting2023model,hong2016probabilistic}. In recent years disease outbreak forecast hubs require that all forecasts submitted by outside participants be represented by  several predictive intervals or quantiles. This standardized representation allows for straightforward forecast scoring and ensemble building \cite[]{mathis2024evaluation,mathis2023flusight,Cramer2022-hub-dataset,cramer2022evaluation,sherratt2023predictive,bracher2021evaluating}.

A set of quantiles may provide distributional information for an event, but it is not as informative as a distribution defined by a cumulative distribution function (CDF), the inverse-CDF or quantile function QF, or a probability density/mass function (PDF), and quantiles provide no distribution tail information --information for values below the smallest quantile or above the largest quantile. Another drawback for using quantiles to define a distribution is that many tools for evaluating or scoring continuous distributions require a CDF or PDF
% \jarad{what does 'fully defined' mean?} 
\cite[]{gneiting2007strictly,gneiting2014probabilistic}. Combining distributions into an ensemble distribution is commonly 
% \jarad{most often? perhaps use commonly}
done by aggregating multiple QFs or 
CDFs/PDFs, the latter being possible only when a CDF/PDF is available \cite[]{gneiting2005calibrated,wang2023forecast}.

Fitting quantiles to recover or estimate a continuous distribution is done in many fields, often for the purpose of evaluating forecasts using rules that require a CDF or PDF \cite[]{simpson2023interpolating, gerding2023evaluating}. Fitting may also be done to be able to combine forecasts using aggregation methods that require CDFs or PDFs \cite[]{gyamerah2020probabilistic,li2019combining,baran2018combining,bogner2017combining,he2016short,gneiting2005calibrated}. An example of fitting quantiles is given in figure \ref{fig:quant_match_example}. The 12 points in the figure are quantiles estimated from a random sample of size 100 from a standard normal distribution for given probabilities. The grey line is the fitted QF of a normal distribution that was estimated by selecting the mean and standard deviation parameters which minimize the least squares distance between the estimated quantiles and the QF.

\begin{figure}[hbt!]
    \centering
    \includegraphics[scale=.5]{Images/fit_quantiles_example.png}
    \caption{Points representing 12 quantiles ($y$-axis) for given probability values ($x$-axis) estimated from a random sample of size 100 from a standard normal distribution. The quantile function (QF) of normal distribution (grey) was fit by selecting the mean and standard deviation parameters which minimize the least squares distance between the quantiles and the QF.}
    \label{fig:quant_match_example}
\end{figure}

Hereafter we refer to estimating a continuous distribution by fitting quantiles as quantile matching (QM).
\cite{sgouropoulos2015matching} performed QM by minimizing the mean square difference between quantiles of a response variable and a linear combination of quantiles of covariates. Selecting distribution parameters which minimize the mean square error
% \jarad{is least squares different than minimizing the mean square difference?}
between quantiles and a QF is a common QM method \cite[]{dilger2022distributions,li2019combining, belgorodski2017quantilemse}, 
% \jarad{Describe, briefly, what these methods do rather than just listing them.}
kernel density estimation and spline interpolation are common nonparametric QM methods \cite[]{gerding2023evaluating,gyamerah2020probabilistic,he2016short}, and \cite{keelin2016metalog} introduced a semiparametric method based on defining a flexible QF to fit quantiles. An issue shared by these methods is that they do not adequately account for the uncertainty inherent from the estimation of quantiles.  \cite{nirwan2020bayesian} introduced a QM model based on the definition of sample quantiles as order statistics and formulated
% \jarad{using as?}
the model likelihood as the joint PDF of multiple order statistics. This model provides exact inference for quantile uncertainty, but relies on the CDF and PDF of the distribution, making it less practical for fitting quantile defined distributions which have a QF but for which the CDF and PDF either do not exist in closed form or are difficult to evaluate \cite[]{perepolkin2023tenets,joiner1971some,tukey1960practical}.

In this manuscript we introduce a novel model, the quantile Gaussian process (QGP), used specifically for QM. The context for using the QGP is that the available data is a set of quantiles estimated for a given set of probabilities.
The QGP relies on established theory underlying the relationship between the QF of a continuous distribution and sample quantiles \cite[]{parzen2004quantile, gilchrist2000statistical,hyndman1996sample,walker1968note,cramer1951mathematical}. The QGP can provide accurate asymptotic inference of uncertainty of quantiles, and it works well for QM for quantile defined distributions. 
In section \ref{sec:quant_func} the QF of a continuous random variable and sample quantiles are each defined, and important properties and asymptotic theory are reviewed. In section \ref{sec:qgp_model} the QGP model for QM is introduced. Section \ref{sec:simulation_analyses} contains simulation studies illustrating the QGP's ability to make parameter inference and match the distribution from which the quantiles where estimated. Section \ref{sec:qm_analysis} contains an application of QM using the QGP model for quantile forecasts of the 2023-24 United States Centers for Disease Control (CDC) flu forecasting competition, also known as FluSight. Section \ref{sec:qm_conclusion} concludes this manuscript with a short summary and ideas for further development in QM modeling.

@@

\begin{definition}
    \label{def:qf}
    For a cumulative distribution function $F: \vmathbb{R} \rightarrow [0,1]$, the quantile function $Q: [0,1] \rightarrow \vmathbb{R}$ is defined as

    \begin{equation*}
    \label{eq:qf}
        Q(p) = F^{-1}(p) := \emph{inf}\{x \in \vmathbb{R} : F(x) \geq p\}, \quad p \in [0,1]
    \end{equation*}
    An alternative definition not in terms of the cumulative distribution function is a function $Q: [0,1] \rightarrow \vmathbb{R} \cup \{-\infty, \infty\}$, which is nondecreasing and left-continuous. 
\end{definition}
% \jarad{the latter does not sound like a definition to me, but instead lists two properties.}

For most statistical applications, it is useful to define the QF as the inverse of the CDF. However, it is sometimes the case that defining a distribution by its QF first is advantageous, for instance when a distribution is defined by a QF but has no closed form CDF or when simple QFs may be aggregated to define more complex distributions \cite[]{perepolkin2023tenets, gasthaus2019probabilistic, alvarez2023quantile}. This section continues with important properties of the QF and functions including the location-scale property, the quantile density function, the probability integral transform, and examples of quantile defined distributions.
%(should the last reference be included??).

\subsubsection{Location-scale property}
For a random variable with PDF $f_0(x)$ and QF $Q_0(p)$ for $p\in [0,1]$, a location-scale family takes the PDF form $(1/\sigma)f_0((x - \mu)/\sigma)$ where the location parameter $\mu \in (-\infty, \infty)$ and the scale parameter $\sigma \in (0, \infty)$ \cite[]{casella2002statistical}. The QF of the location-scale family then takes the form $\mu + \sigma Q_0(p)$ \cite[]{parzen2004quantile}. By the addition and multiplication rules from section 3.2 of \cite{gilchrist2000statistical}, 
%(should I include a table listing properties somehwere??), 
this is a valid QF. \cite{gilchrist2000statistical} outlines the standardization rule which states that if the random variable $Y$ with QF $Q_Y(p)$ has a standard distribution, that is the random variable $Y$ is centered (by a mean or median) at 0 and has a scale of 1, then a random variable with quantile function $Q_{T(Y)}(p) = \mu + \sigma Q_Y(p)$ is centered at $\mu$ and has scale $\sigma$.

The linear form for a QF is already inherent in some distributions families such as the normal and logistic distribution families, with respective QFs $\mu + \sigma \Phi^{-1} (p)$ and $\mu + \sigma \text{log}(p/(1 - p))$. For the normal distribution, $\Phi^{-1}(p)$ is the QF of a standard normal distribution and $\mu$ and $\sigma$ are the mean and standard deviation of the location-scale transformed normal distribution. For the logistic distribution $\text{log}(p/(1-p))$ is the QF of a logistic distribution with mean 0 and scale 1, and $\mu$ and $\sigma$ become the mean and scale parameters of the location-scale transformed logistic distribution. The linear form is convenient for quantile regression modeling and is key to the development of the metalog distribution family which was designed specifically for modeling quantiles \cite[]{keelin2016metalog}. 
% Two examples of quantile defined distributions  are give further in this section.

@@

Besides the CDF, PDF, and QF a  fourth infrequently mentioned defining function of a continuous random variable is the quantile density function (QDF). As the PDF is the derivative of the CDF, the QDF $q:[0,1] \rightarrow \vmathbb{R}$ is the derivative of the QF defined as
\[
    q(p) = \frac{dQ(p)}{dp}.
\]
If $Q(p)$ is the inverse of a CDF $F(x)$, a calculus result shows that the reciprocal QDF $[q(p)]^{-1} = f(Q(p))$, or the PDF evaluated at the QF at $p$ \cite[]{perepolkin2023tenets, gilchrist2000statistical}. The QDF is important for establishing the CLTs in section \ref{sec:samp_quants} and for the QGP models introduced in section \ref{sec:qgp_model}.

@@

\begin{equation}
    \label{eq:mldq}
    Q_{\textbf{a}}(p) = a_1 + a_3(p - 0.5) + a_2\text{log}\left(\frac{p}{1-p} \right)
\end{equation}
These two distributions will be used in section \ref{sec:simulation_analyses} for comparing different quantile matching methods.

@@

\begin{definition}
For $n$ independent observations $Y_1, ..., Y_n$ from a distribution with corresponding order statistics $Y_{(1)}, ..., Y_{(n)}$, the sample quantile function $\hat{Q}_n(p)$ may be defined as
    \label{def:sqf}
    \begin{equation}
        \label{eq:quant_df}
        \hat{Q}_n(p) = (1- \gamma)Y_{(j)} + \gamma Y_{(j + 1)}
    \end{equation}
    where 
    \[\frac{j - m}{n} \leq p < \frac{j - m + 1}{n}
    \]
    for some $m \in \vmathbb{R}$ and $0 \leq \gamma \leq 1$ where $\gamma$ is some function of $j = \lfloor pn + m \rfloor$,  $g = pn + m - j$, and $\lfloor \cdot \rfloor$ is the floor function.
\end{definition}

@@

The asymptotic distribution of a set of sample quantiles is given in (\ref{eq:qclt}) and is the basis for the QGP model for QM in section \ref{sec:qgp_model}. For independent draws $Y_1, ..., Y_n$ from a continuous random variable with CDF $F$, PDF $f$, and QF $Q$, and with a set of sample quantiles calculated by (\ref{eq:quant_df}), theorem \ref{thm:qclt} holds. Note that in (\ref{eq:qclt}), $N_K(\cdot, \cdot)$ refers to a multivariate normal distribution with $K$ dimensions, and in (\ref{eq:qclt_covariance}) the operator $a \wedge b$ is the minimum of the values $a$ and $b$.

@@

We consider the situation where one is given a set of data including a vector of $K$ probabilities $\boldsymbol{p} = (p_1, ..., p_K)$ and a vector of estimated quantiles at the given probability levels $\hat{\boldsymbol{Q}}_n(\boldsymbol{p}) = (\hat{Q}_n(p_1), ..., \hat{Q}_n(p_K))$. 
The set of quantiles may provide useful information about a distribution, but the information is limited and one may desire a more complete distribution.
The QDP model for QM and estimating a continuous distribution based on the CLT in (\ref{eq:qclt}) is in (\ref{eq:qgp}).

\begin{equation}
    \label{eq:qgp}
\hat{\boldsymbol{Q}}_n(\boldsymbol{p}) \sim N_K(\boldsymbol{Q}_{\theta}(\boldsymbol{p}), n^{-1} \boldsymbol{\mathcal{K}}_{\theta})
\end{equation}
Here $\theta$ is an unknown parameter to be estimated from the data. The covariance matrix $\boldsymbol{\mathcal{K}}_{\theta}$ has entries from the covariance function $\kappa_{\theta}(\cdot, \cdot)$ where 
\[
\kappa_{\theta}(p, p') = \frac{p\wedge p' - p p'}{f_{\theta}(Q_{\theta}(p)) f_{\theta}(Q_{\theta}(p'))}, \quad p, p' \in (0,1)
\]
Here the operator $a\wedge b$ indicates the minimum value between $a$ and $b$.

@@

The KDE method treats given quantiles as if they constitute a random sample from a distribution and applies kernel smoothing to estimate a density function. KDE requires selecting a kernel function, often a Gaussian kernel is chosen, and applying that function to each draw in a sample. \cite{gyamerah2020probabilistic} applied KDE smoothing with an Epanechnikov kernel to quantiles estimated via three different machine learning methods used for predicting crop yield. They then combined the estimated densities into an ensemble prediction. \cite{he2016short} also use KDE to estimate density forecasts from quantiles to produce energy forecasts. We use the \texttt{evmix} package for performing QM using KDE \cite[]{yang2018kde}.

One of the more common methods for QM is to select a CDF or QF function and then select model parameters which give a least squares fit to the estimated quantiles as done by \cite{li2019combining}. 
% and \cite{wadsworth2023mixture}. 
The \texttt{R} package \texttt{rriskDistributions} performs this least squares fit for some standard distributions \cite[]{belgorodski2017quantilemse}. Independent random error may be included to the best fit QF allowing for estimation of the variability of the quantiles \cite[]{nirwan2020bayesian}. In section \ref{sec:simulation_analyses}, we analyze model (\ref{eq:ind_quantile}) as a proxy for the IND model to compare with other methods. 
\begin{equation}
    \label{eq:ind_quantile}
    F_{\theta}(\hat{Q}(p_k)) \overset{ind}{\sim} N(p_k, \sigma_{\rho}^2)
\end{equation}
Here $\hat{Q}_k$ is the estimated sample quantile at probability $p_k$. The major difference between model (\ref{eq:ind_quantile}) and model (\ref{eq:pit_qgp}) is that the error for each quantile in a set of quantiles is considered independent, and the effects of sample size and correlation suggested by the CLT in (\ref{eq:tqclt}) are ignored. Note that the parameter $\sigma_{\rho}$ is not a part of $\theta$, the parameter of the ``true" distribution estimated by QM but instead is meant to capture independent error among the sample quantiles.

>>>>

%% The amssymb package provides various useful mathematical symbols
% \usepackage{amssymb}
%% The amsmath package provides various useful equation environments.
% \usepackage{amsmath}
\usepackage{amsmath,amsthm,amssymb,scrextend,mathtools}
\usepackage{hyperref}
% \usepackage{newtxmath} %for making indicator function
% \RequirePackage[colorlinks,citecolor=blue,urlcolor=blue,backref=page,backref=page]{hyperref}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\usepackage{natbib}
\usepackage{lmodern} %%helps deal with font size problems
\usepackage{dsfont}
\newcommand{\1}[1]{\mathds{1}\left[#1\right]}

@@

%% Abstract
\begin{abstract}
A set of probabilities along with corresponding quantiles are often used to 
define predictive distributions or probabilistic forecasts. These quantile 
predictions offer easily interpreted uncertainty of an event, 
and quantiles are generally straightforward to estimate using standard 
statistical and machine learning methods. However, compared to a distribution 
defined by a probability density or cumulative distribution function, a set of 
quantiles has less distributional information. When given estimated 
quantiles it may be desirable to estimate a fully defined continuous 
distribution function, and many researchers do so to make evaluation or 
ensemble modeling simpler. Most existing methods for fitting a distribution to 
quantiles lack accurate representation of the inherent uncertainty from 
quantile estimation or are limited in their applications. In this manuscript we 
present a Gaussian process model, the quantile Gaussian process, which is based 
on established theory of quantile functions and sample quantiles, to construct 
a probability distribution given estimated quantiles. A Bayesian application of 
the quantile Gaussian process is evaluated for parameter inference, and 
distribution approximation in simulation studies and the quantile Gaussian 
process is used to approximate the distributions of quantile forecasts from the 
2023-24 US Centers for Disease Control collaborative flu forecasting 
initiative. The simulation studies and data analysis show that the quantile 
Gaussian process leads to accurate inference on model parameters, estimation of 
a continuous distribution, and uncertainty quantification of sample quantiles.
\end{abstract}

@@

The use of quantiles in statistical modeling and in reporting inferential or 
predictive uncertainty is widespread, and reporting several quantiles or 
predictive intervals is common for probabilistic forecasting 
\cite[]{gneiting2023model}. Quantiles can be easier to interpret than 
statistical model parameters and are often used to define confidence or 
prediction intervals. Thus with multiple quantiles for a particular outcome, 
one has a measure of uncertainty.
Estimating quantiles in the presence of covariates via 
quantile regression is a common
statistical and machine learning strategy, and where parametric models are 
complicated or nonexistent quantiles may be easier to estimate via machine 
learning methods \cite[]{martin2022direct,chung2021beyond, koenker2017quantile, 
koenker1978regression}. In estimating multiple quantiles, quantile regression 
often provides a tradeoff with parametric modeling in that for some problems 
quantile regression may be easier to perform but that predicted quantiles lack 
the detailed information of a fully defined predictive distribution 
\cite[]{pohle2020murphy}. 
% Quantile regression may also lead to other issues 
% such as quantile crossing where an estimated quantile may be smaller than 
% another estimated quantile with for a lower probability level 
% \cite[]{chernozhukov2010quantile,he1997quantile}.
% Data is often reported as quantiles, 
Perhaps, for data privacy reasons, data quantiles are often reported as 
summaries of the data.
% along with predictive distributions. 
% \jarad{Predictive distributions for what?}
Census or medical data, which can be very large or personal to the subjects, 
may be published as summary or aggregate data including percentiles (quantiles) 
and medians 
\cite[]{simpson2023interpolating,cdc2022growthcharts,nirwan2020bayesian}. In 
collaborative forecast initiatives and competitions, probabilistic forecasts 
are often submitted as a set of predictive quantiles 
% for given probabilities 
\cite[]{gneiting2023model,hong2016probabilistic}. In recent years disease 
outbreak forecast hubs require that all forecasts submitted by outside 
participants be represented by  several predictive intervals or quantiles. 
This standardized representation allows for straightforward forecast scoring 
and ensemble building 
\cite[]{mathis2024evaluation,mathis2023flusight,Cramer2022-hub-dataset,
cramer2022evaluation,sherratt2023predictive,bracher2021evaluating}.

A set of quantiles may provide distributional information for an event, but it 
is not as informative as a distribution defined by a cumulative distribution 
function (CDF), the inverse-CDF or quantile function QF, or a probability 
density/mass function (PDF), and quantiles provide no distribution tail 
information --information for values below the smallest quantile or above the 
largest quantile. Another drawback for using quantiles to define a distribution 
is that many tools for evaluating or scoring continuous distributions require 
a CDF or PDF
% \jarad{what does 'fully defined' mean?} 
\cite[]{gneiting2007strictly,gneiting2014probabilistic}. Likewise combining 
distributions into an ensemble distribution is commonly 
% \jarad{most often? perhaps use commonly}
done by aggregating multiple QFs or 
CDFs/PDFs, the latter being possible only when a CDF/PDF is available 
\cite[]{gneiting2005calibrated,wang2023forecast}.

Fitting quantiles to recover or estimate a continuous distribution is done in 
many fields, often for the purpose of evaluating forecasts using rules that 
require a CDF or PDF \cite[]{simpson2023interpolating, gerding2023evaluating}. 
Fitting may also be done to allow combining multiple forecasts using aggregation 
methods that require CDFs or PDFs 
\cite[]{gyamerah2020probabilistic,li2019combining,baran2018combining,
bogner2017combining,he2016short,gneiting2005calibrated}. 
An example of fitting quantiles is given in figure 
\ref{fig:quant_match_example}. The 12 points in the figure are quantiles 
estimated from a random sample of size 100 from a standard normal distribution 
for given probabilities. The quantile function of the standard normal 
distribution is represented by the solid grey line in the plot. 
The dashed grey line is the fitted QF of a normal 
distribution that was estimated by selecting the mean and standard deviation 
parameters which minimize the least squares distance between the estimated 
quantiles and the QF.

\begin{figure}[hbt!]
    \centering
    \includegraphics[scale=.5]{Images/fit_quantiles_example.png}
    \caption{Points representing 12 quantiles ($y$-axis) for given probability 
    values ($x$-axis) estimated from a random sample of size 100 from a 
    standard normal distribution with the true quantile function (QF) 
    (solid grey). The estimated QF (dashed grey) was fit by selecting the mean 
    and standard deviation 
    parameters which minimize the least squares distance between the quantiles 
    and the QF.}
    \label{fig:quant_match_example}
\end{figure}

Hereafter we refer to estimating a continuous distribution by fitting 
quantiles as quantile matching (QM).
\cite{sgouropoulos2015matching} performed QM by minimizing the mean square 
difference between quantiles of a response variable and a linear combination 
of quantiles of covariates. Selecting distribution parameters which minimize 
the mean square error
% \jarad{is least squares different than minimizing the mean square difference?}
between quantiles and a QF is a common QM method 
\cite[]{dilger2022distributions,li2019combining, belgorodski2017quantilemse}, 
% \jarad{Describe, briefly, what these methods do rather than just listing them.}
kernel density estimation and spline interpolation are common nonparametric 
QM methods 
\cite[]{gerding2023evaluating,gyamerah2020probabilistic,he2016short}, 
and \cite{keelin2016metalog} introduced a semiparametric method based on 
defining a flexible QF to fit quantiles. An issue shared by these methods is 
that they do not adequately account for the uncertainty inherent from the 
estimation of quantiles.  \cite{nirwan2020bayesian} introduced a QM model 
based on the definition of sample quantiles as order statistics and formulated
% \jarad{using as?}
a model likelihood for the quantiles as the joint PDF of multiple order 
statistics. This model 
provides exact inference for quantile uncertainty, but relies on the CDF and 
PDF of the distribution, making it less practical for fitting quantile defined 
distributions which have a QF but for which the CDF and PDF either do not exist 
in closed form or are difficult to evaluate 
\cite[]{perepolkin2023tenets,joiner1971some,tukey1960practical}.

In this manuscript we introduce a novel model, the quantile Gaussian process 
(QGP), used specifically for QM. The context for using the QGP is that the 
available data is a set of quantiles estimated for a given set of 
probabilities.
The QGP relies on established theory underlying the relationship between the 
QF of a continuous distribution and sample quantiles 
\cite[]{parzen2004quantile, gilchrist2000statistical,hyndman1996sample,
walker1968note,cramer1951mathematical}. 
The QGP can provide accurate asymptotic inference of the quantile uncertainty, 
and it works well for QM when modeling quantile defined distributions. 
In section \ref{sec:quant_func} the QF of a continuous random variable and 
sample quantiles are each defined, and important properties and asymptotic 
theory are reviewed. In section \ref{sec:qgp_model} the QGP model for QM is 
introduced. Section \ref{sec:simulation_analyses} contains simulation studies 
illustrating the QGP's ability to make parameter inference and match the 
distribution from which the quantiles where estimated. Section 
\ref{sec:qm_analysis} contains an application of QM using the QGP model for 
quantile forecasts of the 2023-24 United States Centers for Disease Control and
Prevention
(CDC) flu forecasting competition, also known as FluSight. Section 
\ref{sec:qm_conclusion} concludes this manuscript with a short summary and 
ideas for further development in QM modeling.

@@

\begin{definition}
    \label{def:qf}
    For a cumulative distribution function $F: \mathbb{R} \rightarrow 
    \left[ 0,1 \right]$, 
    the quantile function $Q: \left[0,1\right] 
    \rightarrow \mathbb{R}$ is defined as

    \begin{equation*}
    \label{eq:qf}
        Q\left(p\right) = F^{-1}\left(p\right) := 
        \emph{inf}\{x \in \mathbb{R} : F\left(x\right) \geq p\}, 
        \quad p \in \left[0,1\right]
    \end{equation*}
    An alternative definition not in terms of the cumulative distribution 
    function is a function 
    $Q: \left[0,1\right] \rightarrow \mathbb{R} \cup \{-\infty, \infty\}$, which is 
    nondecreasing and left-continuous. 
\end{definition}
% \jarad{the latter does not sound like a definition to me, but instead lists two properties.}

For most statistical applications, it is useful to define the QF as the inverse 
of the CDF. However, it is sometimes the case that defining a distribution by 
a QF rather than a CDF is advantageous. For instance when a distribution is 
defined by a 
QF but has no closed form CDF or when simple QFs may be aggregated to define 
more complex distributions 
\cite[]{perepolkin2023tenets, gasthaus2019probabilistic, alvarez2023quantile}. 
This section continues with important properties of the QF and functions 
including the location-scale property, the quantile density function, the 
probability integral transform, and examples of quantile defined distributions.
%(should the last reference be included??).

\subsubsection{Location-scale property}
For a random variable with PDF $f_0\left(x\right)$ and 
QF $Q_0\left(p\right)$ for $p \in \left[0,1\right]$, a 
location-scale family takes the PDF form 
$\left(1/\sigma\right)f_0\left(\left(x - \mu\right)/\sigma\right)$ 
with location parameter 
$\mu \in \left( -\infty, \infty \right)$ and scale parameter 
$\sigma \in \left(0, \infty \right)$ \cite[]{casella2002statistical}. 
The QF of the 
location-scale family then takes the form $\mu + \sigma Q_0\left(p \right)$ 
\cite[]{parzen2004quantile}. By the addition and multiplication rules from 
section 3.2 of \cite{gilchrist2000statistical}, 
%(should I include a table listing properties somehwere??), 
this is a valid QF. \cite{gilchrist2000statistical} outlines the 
standardization rule which states that if the random variable $Y$ with QF 
$Q_Y\left(p \right)$ has a standard distribution, that is the 
random variable $Y$ is 
centered (by a mean or median) at 0 and has a scale of 1, then a random 
variable with quantile function 
$Q_{T\left(Y \right)}\left(p \right) = \mu + \sigma Q_Y\left(p \right)$ is 
centered at $\mu$ and has scale $\sigma$.

The linear form for a QF is already inherent in some distributions families 
such as the normal and logistic distribution families, with respective QFs 
$\mu + \sigma \Phi^{-1} \left(p \right)$ and 
$\mu + \sigma \text{log}\left(p/ \left(1 - p \right) \right)$. For the
normal distribution, 
$\Phi^{-1}\left(p \right)$ is the QF of a standard normal distribution 
and $\mu$ and $\sigma$ are the mean and standard deviation of the 
location-scale transformed normal distribution. For the logistic distribution 
$\text{log}\left(p/\left(1-p \right) \right)$ 
is the QF of a logistic distribution with mean 0 and 
scale 1, and $\mu$ and $\sigma$ become the mean and scale parameters of the 
location-scale transformed logistic distribution. The linear form is convenient 
for quantile regression modeling and is key to the development of the metalog 
distribution family which was designed specifically for modeling quantiles 
\cite[]{keelin2016metalog}. 
% Two examples of quantile defined distributions  are give further in this section.

@@

Besides the CDF, PDF, and QF a  fourth infrequently mentioned defining function 
of a continuous random variable is the quantile density function (QDF). As the 
PDF is the derivative of the CDF, the QDF $q:[0,1] \rightarrow \mathbb{R}$ is 
the derivative of the QF defined as
\[
    q(p) = \frac{dQ(p)}{dp}.
\]
If $Q(p)$ is the inverse of a CDF $F(x)$, a calculus result shows that the 
reciprocal QDF $[q(p)]^{-1} = f(Q(p))$, or the PDF evaluated at the QF 
evaluated at $p$ 
\cite[]{perepolkin2023tenets, gilchrist2000statistical}. The QDF is important 
for establishing the CLTs in section \ref{sec:samp_quants} and for the QGP 
models introduced in section \ref{sec:qgp_model}.

@@

\begin{equation}
    \label{eq:mldq}
    Q_{\textbf{a}}(p) = a_1 + a_3(p - 0.5) + a_2\text{log}\left(\frac{p}{1-p} \right)
\end{equation}
These two distributions are used in section \ref{sec:simulation_analyses} 
for comparing different quantile matching methods.

@@

\begin{definition}
For $n$ independent observations $Y_1, ..., Y_n$ from a distribution with 
corresponding order statistics $Y_{(1)}, ..., Y_{(n)}$, the sample quantile 
function $\hat{Q}_n(p)$ may be defined as
    \label{def:sqf}
    \begin{equation}
        \label{eq:quant_df}
        \hat{Q}_n(p) = (1- \gamma)Y_{(j)} + \gamma Y_{(j + 1)}
    \end{equation}
    where 
    \[\frac{j - m}{n} \leq p < \frac{j - m + 1}{n}
    \]
    for some $m \in \mathbb{R}$ and $0 \leq \gamma \leq 1$ where 
    $\gamma$ is some function of $j = \lfloor pn + m \rfloor$,  
    $g = pn + m - j$, and $\lfloor \cdot \rfloor$ is the floor function.
\end{definition}

@@

The asymptotic distribution of a set of sample quantiles is given in 
(\ref{eq:qclt}), and it is the basis for the QGP model for QM in section 
\ref{sec:qgp_model}. For independent draws $Y_1, ..., Y_n$ from a continuous 
random variable with CDF $F$, PDF $f$, and QF $Q$, and with a set of sample 
quantiles calculated by (\ref{eq:quant_df}), theorem \ref{thm:qclt} holds. 
Note that in (\ref{eq:qclt}), $N_K(\cdot, \cdot)$ refers to a multivariate 
normal distribution with $K$ dimensions, and in (\ref{eq:qclt_covariance}) the 
operator $a \wedge b$ is the minimum of the values $a$ and $b$.

@@

We consider the situation where one is given a set of data including a vector 
of $K$ probabilities $\boldsymbol{p} = (p_1, ..., p_K)$ and a vector of 
estimated quantiles at the given probability levels 
$\hat{\boldsymbol{Q}}_n(\boldsymbol{p}) = (\hat{Q}_n(p_1), ..., \hat{Q}_n(p_K))$. 
The set of quantiles may provide useful information about a distribution, but 
the information is limited and one may desire a more complete distribution.
The QGP model for QM and estimating a continuous distribution based on the CLT 
in (\ref{eq:qclt}) is in (\ref{eq:qgp}).

\begin{equation}
    \label{eq:qgp}
\hat{\boldsymbol{Q}}_n(\boldsymbol{p}) \sim N_K(\boldsymbol{Q}_{\theta}(\boldsymbol{p}), n^{-1} \boldsymbol{\mathcal{K}}_{\theta})
\end{equation}
Here $\theta$ is an unknown parameter to be estimated from the data. The 
covariance matrix $\boldsymbol{\mathcal{K}}_{\theta}$ has entries from the 
covariance function $\kappa_{\theta}(\cdot, \cdot)$ where 
\[
\kappa_{\theta}(p, p') = \frac{p\wedge p' - p p'}{f_{\theta}(Q_{\theta}(p)) f_{\theta}(Q_{\theta}(p'))}, \quad p, p' \in (0,1)
\]

@@

The KDE method treats given quantiles as if they constitute a random sample 
from a distribution and applies kernel smoothing to estimate a density 
function. KDE requires selecting a kernel function, often a Gaussian kernel is 
chosen, and applying that function to each draw of a sample. 
\cite{gyamerah2020probabilistic} applied KDE smoothing with an Epanechnikov 
kernel to quantiles estimated via three different machine learning methods 
used for predicting crop yield. They then combined the estimated densities 
into an ensemble prediction. \cite{he2016short} also use KDE to estimate 
density forecasts from quantiles to produce energy forecasts. We use the 
\texttt{evmix} package for performing QM using KDE \cite[]{yang2018kde}.

One of the more common methods for QM is to select a CDF or QF function and 
then select model parameters which give a least squares fit to the estimated 
quantiles as done by \cite{li2019combining}. 
% and \cite{wadsworth2023mixture}. 
The \texttt{R} package \texttt{rriskDistributions} performs this least squares 
fit for some standard distributions \cite[]{belgorodski2017quantilemse}. 
Independent random error may be included to the best fit QF allowing for 
estimation of the variability of the quantiles \cite[]{nirwan2020bayesian}. 
In section \ref{sec:simulation_analyses}, we analyze model 
(\ref{eq:ind_quantile}) as a proxy for the IND model to compare with 
other methods. 
\begin{equation}
    \label{eq:ind_quantile}
    F_{\theta}(\hat{Q}(p_k)) \overset{ind}{\sim} N(p_k, \sigma_{\rho}^2)
\end{equation}
Here $\hat{Q}_k$ is the estimated sample quantile at probability $p_k$. 
The major difference between model (\ref{eq:ind_quantile}) and model 
(\ref{eq:pit_qgp}) is that the error for each quantile in a set of quantiles is 
considered independent thus ignoring the effects of sample size and correlation 
suggested by the CLT in (\ref{eq:tqclt}). Note that the parameter 
$\sigma_{\rho}$ is not a part of $\theta$, the parameter of the ``true`` 
distribution estimated by QM but instead is meant to capture independent 
error among the sample quantiles.
